The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-04-12 paper.tex
Restructure.

\section{Developing ANNO}
%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
%First, \emph{task ontology} is created as a conceptual model of \aw{}.
%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
\todo{figure out how to best separate ontology and software}
The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.

@@

% todo: do we have an ontology engineering methodology?
% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
% TODO: ontology language, expressivity
% TODO: reuse of ontological entities
% TODO: clear validation mechanism
% TODO: top-level ontology?
% @Alex ist das CSV->OWL Script irgendwo publiziert, kann man das zitieren?
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}
The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
\todo{können wir es auf OWL 2 DL eingrenzen?}
% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.\todo{explain}

\subsection{Process for creating definitions and measurements}
For the definitions and measurement, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
Notably, Latin or ancient Greek terms were missing from the English-language literature.
The FMA also contains only English terms.
For this reason, these were also searched out with their synonyms.
Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
Where there were contradictions in the literature, the information was highlighted, reconciled, and logically checked.
For the osteometric landmarks, the abbreviations and their synonyms of the three languages were also provided.
Furthermore, for these as well as for the measured sections, the sources were divided into parent source, core source and occurrence source.
The parent source designates the original source of the landmark.
Under the core source, those were listed where the landmarks are listed and sampled by default.
Occurrence source represents the sources where the landmark occurs for measurements.
For redefinitions of landmarks, a meaningful Latin name had to be selected.
Requirements for this were an included position description (e.g. Punctum superioris capitis femoris as superior located point of the Caput femoris) necessary.
For this, the examination of the Latin grammar was relevant.
For the measurement sections, the type of measurement (e.g., distance measurement) and the measurement instrument were taken in addition to the name of the measurement section.
The subsequently created visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric ones.

\subsection{Functions}
The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
Names were assigned by the number included in each study, the authors, and the year of publication.
In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.

\section{Description of ANNO}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNO-core ontology}\label{fig:core}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.3\textwidth]{img/modules.pdf}
\caption{The six modules of ANNO}\label{fig:modules}
\end{figure}

@@

\subsection{Teeth}
\subsection{Measurements}
For measurements consist of references to sections, the start point, midpoint, end point, and short definitions.

\subsection{Functions}
Functions are categorized into sex determination and regress functions for body height estimation.
RDF is not optimized for mathematical formulas so we model those as literals.

@@

TODO: Modellierung der Anthropologie-Domäne (u.a. Motivation, related work, was ist neu).
TODO: Dann könnte man Design-Entscheidungen beschreiben (wie und warum wurden bestimmte Sachverhalte so in der Ontologie modelliert).
\section{Publication}
ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
TODO: Bereitstellung der Ontologie für die Community (OLS etc.).

@@

\section{Conclusion and Future Work}
% Basiert auf dem Fazit aus dem Projektbericht
ANNO introduces uniform definitions for the anatomical landmarks of the human skeletal system as well as osteometric landmarks, measurement distances and functions.
basis for anthropological work
The ontology is interlinked with ... and includes sources for all external definitions.
The ontology is used in \aw{} in place of an old, hard-coded format, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and increases interoperability.
% following two sentences are machine translated, improve and rewrite
It provides a basis for future training of ontologists, knowledge managers, further research activities and specialization opportunities in this field.
Furthermore, there is interest of the results in forensic, historical and prehistoric anthropology, pathology and medicine as well as in the field of computer science and especially medical informatics.
%Moreover, the ontology can be extended.

can be extended with osteological terms of description
Arten von Erhebungen, Öffnungen, ...
anthropologische ebene draufmodellieren zb form
an anthropological landmark can have different anthropological properties

>>>>

\section{Ontological Framework}
%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
%First, \emph{task ontology} is created as a conceptual model of \aw{}.
%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
\todo{figure out how to best separate ontology and software}
The \aw{} software is designed according to the three-ontology method~\citep{threeontologymethod}.
This method is based on the interaction of three different types of ontologies: a task ontology (TO), a domain ontology (DO) and a top-level ontology (TLO).
The TO serves as the conceptual model for the software, the DO provides the domain-specific knowledge, while the TLO integrates the TO and the DO and is used as foundation for them.

@@

% take care to prevent the problems noted by reviewer 1 in https://semantic-web-journal.net/content/modelling-digital-health-data-examode-ontology-histopathology
% TODO: ontology language, expressivity
% TODO: reuse of ontological entities
% TODO: clear validation mechanism
% TODO: top-level ontology?

\section{Description of ANNO-core}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/core.pdf}
\caption{ANNO-core ontology}\label{fig:core}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.3\textwidth]{img/modules.pdf}
\caption{The six modules of ANNO}\label{fig:modules}
\end{figure}
\todo{Brauchen wir \cref{fig:modules} noch oder ist sie durch \cref{fig:core} nicht mehr nötig?}

@@

\subsection{Teeth}
\subsection{Measurements}
Measurements consist of references to sections, the start point, midpoint, end point, and short definitions.

\todo{Umschreiben von Modulen auf ANNO-core}
\subsection{Functions}
Functions are categorized into sex determination and regress functions for body height estimation.
RDF is not optimized for mathematical formulas so we model those as literals.

@@

\section{Development of ANNO-domain}

\begin{figure}[h!t]
\includegraphics[width=\textwidth]{img/smog.png}
\caption{Excerpt of the spreadsheet-based input template used by the anthropologists.}\label{fig:smog}
\end{figure}

The domain experts are provided a spreadsheet-based SMOG~\citep{smog} template by the ontologists, see \cref{fig:smog}.
The spreadsheet is transformed to an OWL 2 ontology consisting of a taxonomy, annotations and some simple axioms on the basis of property restrictions.
\todo{können wir es auf OWL 2 DL eingrenzen?}

% ist das richtig übersetzt?
Sources are categorized into core, root, and occurrence sources.\todo{explain}

\subsection{Process for creating definitions and measurements}
For the definitions and measurement, a minimum amount of English-language and mostly German-language literature was compared in order to develop the definitions from their information.
Notably, Latin or ancient Greek terms were missing from the English-language literature.
The FMA also contains only English terms.
For this reason, these were also searched out with their synonyms.
Overall, the name of the landmark or survey route had to be noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
Where there were contradictions in the literature, the information was highlighted, reconciled, and logically checked.
For the osteometric landmarks, the abbreviations and their synonyms of the three languages were also provided.
Furthermore, for these as well as for the measured sections, the sources were divided into parent source, core source and occurrence source.
The parent source designates the original source of the landmark.
Under the core source, those were listed where the landmarks are listed and sampled by default.
Occurrence source represents the sources where the landmark occurs for measurements.
For redefinitions of landmarks, a meaningful Latin name had to be selected.
Requirements for this were an included position description (e.g. Punctum superioris capitis femoris as superior located point of the Caput femoris) necessary.
For this, the examination of the Latin grammar was relevant.
For the measurement sections, the type of measurement (e.g., distance measurement) and the measurement instrument were taken in addition to the name of the measurement section.
The subsequently created visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric ones.
\subsection{Functions}
The functions were divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
The goal here was to cover functions for at least one European, African, American, and Asian ethnicity or population.
Names were assigned by the number included in each study, the authors, and the year of publication.
In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.

@@

\section{Conclusion and Future Work}
% Conclusion

% Basiert auf dem Fazit aus dem Projektbericht
ANNO introduces uniform definitions for the anatomical landmarks of the human skeletal system as well as osteometric landmarks, measurement distances and functions.
basis for anthropological work
The ontology is interlinked with ... and includes sources for all external definitions.
The ontology is used in \aw{} in place of an old, hard-coded format, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and increases interoperability.
% following two sentences are machine translated, improve and rewrite
It provides a basis for future training of ontologists, knowledge managers, further research activities and specialization opportunities in this field.
Furthermore, there is interest of the results in forensic, historical and prehistoric anthropology, pathology and medicine as well as in the field of computer science and especially medical informatics.
%Moreover, the ontology can be extended.

ANNO is published on \url{https://ols.imise.uni-leipzig.de} as well as \url{https://annosaxfdm.de/ontology/}.
Note: Currently on \url{https://annosaxfdm.imise.uni-leipzig.de/ontology/}.
TODO: Bereitstellung der Ontologie für die Community (OLS etc.).

% Future Work
can be extended with osteological terms of description
Arten von Erhebungen, Öffnungen, ...
anthropologische ebene draufmodellieren zb form
an anthropological landmark can have different anthropological properties
